---
title: "Generation of signature interpretation matrix from simulated genomes"
output: 
  pdf_document:
    number_sections: true
---

```{r setup, message=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)

options(java.parameters = "-Xmx8000m")
library(data.table)
library(dplyr)
library(ggpubr)
library(ggrepel)
library(sjPlot)
library(tidyverse)
library(rstatix)
library(reshape2)
library(ggplot2)
library(ggthemes)
library(lemon)
library(RColorBrewer)
library(coin)

# PATHWAYS
BASE <- dirname(rstudioapi::getSourceEditorContext()$path)
INPUT_DIR <- file.path(BASE,"data")
OUTPUT_DIR <- file.path(BASE,"output")
dir.create(OUTPUT_DIR, showWarnings = FALSE, recursive = TRUE)

# PLOTS THEME
theme_set(theme_tufte(base_size = 6, base_family = "ArialMT"))
theme_update(text = element_text(size = 6),
             axis.text = element_text(size = 6),
             axis.line = element_line(size = 0.5), 
             axis.ticks = element_line(size = 0.5),
             axis.ticks.length = unit(.1, "cm"))
```

# LOAD DATA
```{r data}
#CIN process simulated per sample 
Simulations_overview <- readRDS(paste0(INPUT_DIR,"/Out_2_Step_3_Overview_simulation_20each_N240_RubenParam2.rds"))
colnames(Simulations_overview)[2] <- "sampleID"

#Normalize matrix
#Sample per components matrix
Simulations_components <- t(read.table(paste0(INPUT_DIR,"/Out_4_Step_3_CxS_Matrix_20each_N240_RubenParam2_noChimeric.txt"), sep="\t", header = T, check.names=FALSE))
#Normalize matrix
Simulations_components[,c(1:22)] <- t(apply(Simulations_components[,c(1:22)],1,function(x){x/sum(x)}))
Simulations_components[,c(23:32)] <- t(apply(Simulations_components[,c(23:32)],1,function(x){x/sum(x)}))
Simulations_components[,c(33:35)] <- t(apply(Simulations_components[,c(33:35)],1,function(x){x/sum(x)}))
Simulations_components[,c(36:40)] <- t(apply(Simulations_components[,c(36:40)],1,function(x){x/sum(x)}))
Simulations_components[,c(41:43)] <- t(apply(Simulations_components[,c(41:43)],1,function(x){x/sum(x)}))
components <- colnames(Simulations_components)

#CNSig exposure per sample
Simulations_signatures <- readRDS(paste0(INPUT_DIR,"/Out_6_Activities_20each_N240_RubenParam2_CxSBefore_SigsBefore.rds"))

#CNSig definitions (weights per component)
SIMSig_definitions <- readRDS(paste0(INPUT_DIR,"/Out_6_Signatures_XK5TRD_normalised_RubenParam2_noChimeric.rds"))
```

# ACTIVITY THRESHOLDS
We calculated signature-specific thresholds to ensure activity levels are true. Activity levels below to the specific threshold are corrected to zero.
```{r}
## Repair names
vNamesRepair = sapply(as.character(rownames(Simulations_signatures)), function(thisName) {
  if(nchar(thisName) == 7) {
    return(substr(thisName, 2, 7))
  } else {
    return(thisName)
  }
})
rownames(Simulations_signatures) <- vNamesRepair

mutprocesses <- c("LST","ecDNA","Chr","WGDearly","WGDlate")
Corrected_Signatures <- c()
for (i in 1:ncol(Simulations_signatures)){
  act <- as.data.frame(cbind(sampleID=rownames(Simulations_signatures),level=Simulations_signatures[,i]))
  
  #get samples with no signature operating
  Simulations_overview$samps <- FALSE
  for (j in 1:nrow(Simulations_overview)){
    if(i<=3 & !str_detect(Simulations_overview$Signature[j], "WGD")){
      Simulations_overview$samps[j] <- TRUE
    }
    if(i<=3 & str_detect(Simulations_overview$Signature[j], "WGD") & str_detect(Simulations_overview$Signature[j], mutprocesses[i])){
      Simulations_overview$samps[j] <- TRUE
    }
    if(i==4 & str_detect(Simulations_overview$Signature[j], "WGDearly")){
      Simulations_overview$samps[j] <- TRUE
    }
    if(i==5 & str_detect(Simulations_overview$Signature[j], "WGDlate")){
      Simulations_overview$samps[j] <- TRUE
    }
  }
  noAct <- Simulations_overview$sampleID[Simulations_overview$samps==FALSE]
  dt <- left_join(act, Simulations_overview, by="sampleID")
  noAct <- dt[dt$samps==FALSE,]
  
  #set threshold in 80%
  noAct$level <- as.numeric(noAct$level)
  noAct <- noAct[order(-noAct$level),]
  thr <- nrow(noAct)-(80*nrow(noAct)/100)
  thr <- noAct[thr,2]
  
  #change to zero these samples with value lower than threhold
  act$level[act$level<=thr]<-0
  Corrected_Signatures<-cbind(Corrected_Signatures,as.numeric(act$level))
}
colnames(Corrected_Signatures)<-colnames(Simulations_signatures)
rownames(Corrected_Signatures)<-rownames(Simulations_signatures)
```

# LINEAR CORRELATIONS
For each signature, we detect components that are positively correlated with activity levels
```{r lm}
CNsigs <- colnames(Corrected_Signatures)
component <- colnames(Simulations_components)

#Sig activity
sigs <- reshape2::melt(Corrected_Signatures)
colnames(sigs) <- c("sampleID", "CNSig", "level")
## Repair names
sigs$sampleID <- as.character(sigs$sampleID)
vNamesRepair = sapply(sigs$sampleID, function(thisName) {
  if(nchar(thisName) == 7) {
    return(substr(thisName, 2, 7))
  } else {
    return(thisName)
  }
})
sigs$sampleID <- vNamesRepair

#Sum-of-posterior
comps <- reshape2::melt(t(Simulations_components))
colnames(comps)<-c("component", "sampleID", "value")

#Join
dat <- left_join(sigs,comps, by="sampleID")

#Linear correlation 
res <- dat[dat$level!=0,] %>% group_by(CNSig,component) %>% do(tidy(lm(level ~ value, data = .)))
res$beta <- res$term=="value"
res <- res[res$beta==TRUE,] #remove intercept values
#Correct pvalues
res$p.adj <- p.adjust(res$p.value, "BH")
```

# MIX INFORMATION IN INPUT MATRIX AND ACTIVITY MATRIX
```{r mix}
# Denormalize components matrix
comps <- Simulations_components
comps[,c(1:22)] <- t(apply(comps[,c(1:22)],1,function(x){x*22}))
comps[,c(23:32)] <- t(apply(comps[,c(23:32)],1,function(x){x*10}))
comps[,c(33:35)] <- t(apply(comps[,c(33:35)],1,function(x){x*3}))
comps[,c(36:40)] <- t(apply(comps[,c(36:40)],1,function(x){x*5}))
comps[,c(41:43)] <- t(apply(comps[,c(41:43)],1,function(x){x*3}))
comps <- reshape2::melt(t(comps))
colnames(comps)<-c("component", "sampleID", "value")
dat <- left_join(sigs,comps, by="sampleID")

# Multiply component weights and activity level per sample
dat$strength <- dat$value*dat$level

# Sum strength of each component per signature
dat <- dat %>% group_by(CNSig,component) %>% summarise(sum=sum(strength))
dat <- left_join(dat, res, by=c("CNSig", "component"))

# Normalize values to one
dat[,"sum"] <- apply(dat[,"sum"],2,function(x){x/sum(x)})
```

# ADD INFORMATION OF SIGNATURE DEFINITIONS MATRIX FOR GENERATING THE INTERPRETATION MATRIX
```{r interpret}
defs <- reshape2::melt(t(SIMSig_definitions))
colnames(defs)<-c("component", "CNSig", "weights")
defs$CNSig <- as.character(defs$CNSig)
defs$CNSig[defs$CNSig=="W3"]<-"HRDLST"
defs$CNSig[defs$CNSig=="W5"]<-"ecDNA"
defs$CNSig[defs$CNSig=="W1"]<-"CHR"
defs$CNSig[defs$CNSig=="W4"]<-"WGDearly"
defs$CNSig[defs$CNSig=="W2"]<-"WGDlate"
dat <- left_join(dat, defs, by=c("CNSig", "component"))

# Normalize values to one
dat[,"weights"] <- apply(dat[,"weights"],2,function(x){x/sum(x)})

# Sum values from both matrices
dat$allSum <- dat$sum + dat$weights
```

# PLOT POSITIVELY CORRELATED COMPONENTS 
```{r plot}
## Prepare data for plotting
# 1. Select samples with significant (pvalue<0.05) and positive (beta>0) correlations
dat$plot <- dat$estimate>=0 #& dat$p.value<0.05

# 2. Mask no-significant
dat$allSum[dat$plot==FALSE]<-0

# 3. Normalize sum per signature & feature
dt <- list()
dt <- lapply(CNsigs, function(sig){
  s <- dat[dat$CNSig==sig,]
  # s[,"allSum"] <- apply(s[,"allSum"],2,function(x){x/sum(x)})
  s[c(1:22),"allSum"] <- apply(s[c(1:22),"allSum"],2,function(x){x/sum(x)})
  s[c(23:32),"allSum"] <- apply(s[c(23:32),"allSum"],2,function(x){x/sum(x)})
  s[c(33:35),"allSum"] <- apply(s[c(33:35),"allSum"],2,function(x){x/sum(x)})
  s[c(36:40),"allSum"] <- apply(s[c(36:40),"allSum"],2,function(x){x/sum(x)})
  s[c(41:43),"allSum"] <- apply(s[c(41:43),"allSum"],2,function(x){x/sum(x)})
  dt[[sig]]<-s
})
names(dt)<-CNsigs
dt <- do.call(rbind, dt)
#dt$allSum[is.nan(dt$allSum)]<-0

# 4. Define categories of variables
dt$component <- factor(dt$component, levels = rev(component))
dt$CNSig <- factor(dt$CNSig, levels = CNsigs)

# 5. Discretize signature interpretation values
breaks = c(0, 1e-11, 0.05, 0.1, 0.5, 1)
# Names of categories
tags = c("0", ">0%", ">5%", ">10%", ">50%")
# Split allSum into somewhat meaningful categories
dt$Disc = cut(dt$allSum, breaks = breaks, include.lowest=TRUE, right=FALSE, labels = tags)
# Colors
# blue gradient
cols = c("0" = "#FFFFFF", ">0%" = "#ece7f2", 
         ">5%" = "#a6bddb", ">10%"= "#3690c0", 
         ">50%"= "#045a8d")

# Plot mean of sum of posterior values per component in the group with high activity
pB = ggplot(dt, aes(y = component, x = CNSig, fill = Disc)) + 
  geom_tile(aes(width = 0.94, height = 0.94)) + 
  theme_tufte(base_family = "", base_size = 16) + 
  theme(legend.position = "bottom", legend.justification = "right", legend.title = element_text(size=12), legend.text = element_text(size=10), 
        axis.line = element_line(size = 0.5), axis.ticks = element_line(size = 0.5), 
        axis.ticks.length = unit(.1, "cm"), plot.margin = unit(c(0, 0, 0, 0), "null"), aspect.ratio = 43/17) + 
  labs(x = "CIN signature", y = "Feature components") + 
  scale_x_discrete(position = "top", guide = guide_axis(angle = 90)) + labs(fill = "Interpretation values\nBeta > 0") +
  scale_fill_manual(values = cols) + 
  guides(fill = guide_legend(nrow = 1)) +
  coord_capped_cart(top = "both", left = "both")

print(pB)
# ## Save output
# cairo_pdf(file.path(OUTPUT_DIR, "Figure_IntMatrix_Simulations_80thThrehold.pdf"), height = 230/25.4, width = 161/25.4)
# print(pB); dev.off()
ggsave(file.path(OUTPUT_DIR, "SuppFig_xx_Simulations_signature_interpretation_vertical.svg"), pB, height = 230/25.4, width = 161/25.4)

# ## Save file
write.table(dt, file=paste0(OUTPUT_DIR,"/InterpretationMatrix_simulations.txt"), sep="\t", col.names = TRUE, quote = FALSE)
```
